---
title: "CaseStady2_Max"
author: "Max Moro"
date: "November 26, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(openxlsx)
library(reshape2)
library(forcats)
library(scales)
library(xgboost)
library(caret)
library(knitr)
```

# Functions 

```{r}
plotTrend = function(data,x,y){
  #function to create a boxplot comparing categorical and numeric columns
  data2=data 
  data2$x=forcats::fct_reorder(data[[x]],data[[y]],.fun=median)
  data2$y=data2[[y]]
  
  ggplot2::ggplot(data=data2,aes(x=x,y=y)) +
    geom_boxplot(aes(fill=x)) +
    coord_flip() +
    xlab(x)+
    ylab(y)+
    ggplot2::ggtitle(paste('Comparing',y,'and',x)) +
    theme_minimal() +
    scale_y_continuous(labels=comma_format(accuracy=1)) + 
    scale_fill_brewer(palette='Blues',guide=F)
}

plotSurvey = function(data,x,y){
  #function to create a text plot comparing survey data
  data2=data 
  data2$x=forcats::fct_reorder(data[[x]],data[[y]],.fun=median)
  data2$y=as.character(data2[[y]])
  data2 = data2 %>% group_by(x,y)  %>%
    summarise(count=n()) %>%
    group_by(x) %>%
    mutate(perc=count/sum(count)) %>%
    ungroup() %>%
    mutate(color = ifelse(y>=3,-perc,perc))
  
  ggplot2::ggplot(data=data2,aes(x=x,y=y,size=perc,label=percent(perc,1))) +
    geom_text(aes(color=color)) +
    coord_flip() +
    xlab(x)+
    ylab(y)+
    ggplot2::ggtitle(paste('Comparing',y,'and',x)) +
    theme_minimal() + 
    scale_size_continuous(guide=F) +
    #scale_color_continuous(guide=F,low='#fcbba1',high='#67000d')
    scale_color_gradient2(guide=F,low='#006d2c',mid='#ffffff',high='#67000d',midpoint=0)
}

plotAttritionBars = function(data,x,sort=T){
  #function to compare attrition across categories
  data2=data 
  data2$x=as.character(data[[x]])
  data2 = data2 %>% 
    group_by(x,Attrition)  %>%
    summarise(count=n()) %>%
    group_by(x) %>%
    mutate(percent=count/sum(count) ) %>%
    filter(Attrition=='Yes') %>%
    ungroup() %>%
    arrange(percent)
  if (sort) data2=    mutate(data2,x=forcats::fct_reorder(x,percent))
  
  ggplot(data=data2,aes(x=x,y=percent,fill=percent))+
    geom_bar(stat='identity') +
    coord_flip()+
    geom_text(aes(label=percent(percent)),size=3,hjust=-.1,color='#a50026')+
    ggplot2::scale_y_continuous(labels=percent_format(accuracy = 1)
                                ,limits=c(0,max(temp$percent+.1))
                                ,name='Attrition Rate') +
    ggplot2::scale_x_discrete(name=x) +
    scale_fill_gradient(low='#fee8c8',high='#a50026',name='Attrition \n Rate',labels=percent_format(accuracy=1))+
    theme_minimal()+
    ylab('Attrition Rate')
}

plotBars = function(data,x,y){
  #function to create box plots
  data2=data 
  data2$x=data2[[x]]
  data2$y=as.factor(data2[[y]])
  
  data2 = data2 %>% 
    group_by(x,y)  %>%
    summarise(count=n()) %>%
    group_by(x) %>%
    mutate(percent=count/sum(count) ) %>%
    ungroup() %>%
    arrange(percent) %>%
    mutate(y=forcats::fct_rev(y))
  
  ggplot2::ggplot(data=data2,aes(x=x,fill=y,y=percent)) +
    geom_col(stat='individual') +
    coord_flip() +
    xlab(x)+
    ylab('% of Workers')+
    ggplot2::ggtitle(paste('Comparing',y,'and',x)) +
    theme_minimal() +
    scale_fill_brewer(palette='Blues',name=y) + 
    scale_y_continuous(label=percent_format(accuracy = 1))
}
```

# Reading  Data Source

```{r read}
dataOri=read.xlsx('data/CaseStudy2-data.xlsx',sheet=1) %>%
  mutate(EmployeeNumber = as.character(EmployeeNumber))

```

# Transformation
```{r transformation}
#checking for na - No NAs in the columns
which(sapply(dataOri,function(x){any(is.na(x))}))
#check for columns with one factor -> 3 columns
oneElement = names(which(sapply(dataOri,function(x){n_distinct(x)==1})))
print(oneElement)
data=dataOri %>% select(-!!oneElement)
```

# Exploratory Data Analysis

## Numerical Columns

### Distribution of Numeric Columns

```{r  fig.height=7, fig.width=8}
#find group with statistical difference

colNum = names(which(sapply(data,is.numeric)))

temp=reshape2::melt(data,id.vars='Attrition',measure.vars=colNum) %>%
  group_by(variable) %>%
  mutate(density = dnorm(value, mean(value), sd(value))) %>% ungroup()

ggplot(data=temp,aes(x=value))+
  geom_histogram(aes(y=..density..),color='gray',bins = 20) +
  geom_line(aes(y = density),   colour = "red") +
  #stat_function(fun = dnorm, args = list(mean = mean(..count..), sd = sd(..count..)),color='red')+
  facet_wrap(~variable,scales='free',ncol=4) +
  theme_minimal()
```

### Applying Log to Some Variables

Following Variable have been transformed with Log function to fit a normal distribution.

```{r  fig.height=5, fig.width=8}
#Testing the Log
colsToLog = c('MonthlyIncome','TotalWorkingYears','YearsAtCompany','YearsInCurrentRole','YearsSinceLastPromotion','YearsWithCurrManager')
temp$log = log(temp$value +.01)

#plot of logged values
ggplot(data=filter(temp,variable %in% colsToLog),aes(x=log))+
  geom_histogram(aes(y=..density..),fill='blue',bins = 20) +
  facet_wrap(~variable,scales='free',ncol=4) +
  ggplot2::ggtitle('Logged Values')+
  theme_minimal()

#Applying the log to the main dataframe
colsLog = paste0(colsToLog,'_log')
dataLog=mutate_at(data,colsToLog,.funs=funs(log=log(. + 0.01))) %>% select(-colsToLog)

```

### Columns with different mean across Attrition 

Following variables have a statisticall significant difference mean between Attrition and Non-Attrition


```{r}
ttestOut=data.frame(col=as.character(),pValue=as.numeric())
colNum = names(which(sapply(dataLog,is.numeric)))
for (col in colNum){
  test=t.test(dataLog[dataLog$Attrition=='Yes',col],dataLog[dataLog$Attrition=='No',col]
              ,alternative = 'two.sided',var.equal=F)
  if(test$p.value<0.01) ttestOut = rbind(ttestOut,data.frame(col=col,pValue=test$p.value))
}

temp=reshape2::melt(dataLog,id.vars='Attrition',measure.vars=ttestOut$col)
ggplot(data=temp,aes(x=Attrition,y=value,fill=Attrition))+
  geom_boxplot() +
  facet_wrap(~variable,scales='free',ncol=4) +
  theme_minimal()  +
  scale_fill_manual(values=c('Yes'='#fc8d59','No'='#91cf60'),guide=F)

```

**List of P-Values for selected variables**

```{r}
kable(arrange(ttestOut,-pValue),digits=4)
```

## Categorical Variables

Distribution of the attrition across categorical values

```{r  fig.height=5, fig.width=10}
#find group with statistical difference
colText = names(which(sapply(select(data,-Attrition,-EmployeeNumber),is.character)))

#make the datast vertical to better manage and chart. Also calculating percentages
temp=reshape2::melt(data,id.vars='Attrition',measure.vars=colText)  %>%
  group_by(variable,value,Attrition) %>%
  summarise(count=n()) %>%
  group_by(variable,value) %>%
  mutate(percent=count/sum(count) ) %>%
  filter(Attrition=='Yes') %>%
  arrange(percent)

#ggplot sort breaks if we have same value across facets.
#to help gplot sort we need to the value+variable insetead of value if there are more than one acorss variables
temp = temp %>% group_by(value)  %>% mutate(cnt = n_distinct(variable)) %>%
  ungroup() %>% mutate(name=ifelse(cnt>1,paste0(value,' (',(substr(variable,1,1)),')'),value))

temp$name=forcats::fct_reorder(temp$name,temp$percent )

ggplot(data=temp,aes(x=name,y=percent,fill=percent))+
  geom_bar(stat='identity') +
  coord_flip()+
  geom_text(aes(label=percent(percent)),size=3,hjust=-.1,color='#a50026')+
  facet_wrap(~variable,scales='free_y',ncol=3) +
  ggplot2::scale_y_continuous(labels=percent_format(accuracy = 1)
                              ,limits=c(0,max(temp$percent+.1))
                              ,name='Attrition Rate') +
  ggplot2::scale_x_discrete(name='Group') +
  scale_fill_gradient(low='#fee8c8',high='#a50026',name='Attrition \n Rate',labels=percent_format(accuracy=1))+
  theme_minimal()
```

# Question 1

*Identify (at least) the top three factors that contribute to turnover.*

## Money Related Factors

### Attrition and Monthly Income
```{r fig.height=3}
plotTrend(data,'Attrition','MonthlyIncome') + scale_fill_manual(values=c('Yes'='#fc8d59','No'='#91cf60'))
```
### Total Working Years vs. Monthly Income

People leaving the company have a lower Income, for > 5 Working Years

```{r fig.height=5}
ggplot(data = data, aes(x = TotalWorkingYears, y = MonthlyIncome, colour = Attrition)) +
  geom_point(alpha=0.8) + 
  geom_smooth(method = 'lm', se = FALSE) + 
  theme_minimal() +
  scale_y_continuous(label=comma_format()) +
  scale_color_manual(values=c('Yes'='#fc8d59','No'='#91cf60'))
```

### Conclusion

The more you are paid, less chance of leaving – no surprise here

## Age and tenure related

### Age

```{r  fig.height=3}
plotTrend(data,'Attrition','Age')  +  scale_fill_manual(values=c('Yes'='#fc8d59','No'='#91cf60'))
```

### Marital Status

```{r  fig.height=4}
plotAttritionBars(data,'MaritalStatus')
```

### Years at Company

```{r fig.height=3}
plotTrend(data,'Attrition','YearsAtCompany')  +  scale_fill_manual(values=c('Yes'='#fc8d59','No'='#91cf60'))
```


### Years in Current role

```{r fig.height=3}
plotTrend(data,'Attrition','YearsInCurrentRole')  +  scale_fill_manual(values=c('Yes'='#fc8d59','No'='#91cf60'))
```

### Years With Current Manager

```{r fig.height=3}
plotTrend(data,'Attrition','YearsWithCurrManager')  +  scale_fill_manual(values=c('Yes'='#fc8d59','No'='#91cf60'))
```

### Conclusion

•	Younger employees and less likely to be marries and are not tied to a job as much. More flexible, hence more attrition. 

•	Younger employees are expected to have less years of service, with company, etc.

•	May need to put in programs to retain younger population.

•	At the same time important to note that we should not introduce age discriminatory practices (company is doing good right now in this respect – see job satisfaction results by age – shows no difference)


## Instrinsic Factors

### Job Satisfaction

```{r fig.height=3}
plotAttritionBars(data,'JobSatisfaction',sort=F)  
```

### Job Involvment

```{r fig.height=3}
plotAttritionBars(data,'JobInvolvement',sort=F)  
```

### Conclusion

People who were more satisfied were not likely to leave 

## Special Mention

**Burnout** due to overtime was another very important factor and needs to be mentioned. Not sure if it fits into a category here, but we should definitely mention it. Conclusion could be that the company needs to do more to control burnout due to overtime. 

```{r  fig.height=3}
plotAttritionBars(data,'OverTime')  
```

# Question 2

Learning about any job role specific trends that may exist in the data set

## Lab Technicians

Research Scientists and Lab technician were almost equal in all aspects – Age, Job Level, Salary, Travel, Overtime etc. However Lab technicians had a 50% more change of leaving. What might be causing this? Recommendation to management - We need to dig in more to get t root cause

```{r fig.height=3}
plotAttritionBars(data,'JobRole')  

dataLabRes = filter(data,JobRole %in% c('Research Scientist','Laboratory Technician'))
plotTrend(dataLabRes,'JobRole','Age') 
plotBars(dataLabRes,'JobRole','JobLevel')
plotBars(dataLabRes,'JobRole','OverTime') 
```

## Human Resources role

HR role seems to be least satisfied in general – does this lead to more attrition (check)

```{r fig.height=3}
plotSurvey(data,'JobRole','JobSatisfaction')
plotAttritionBars(data,'JobRole')  
plotAttritionBars(data,'JobSatisfaction')  
```

## Sales Reps

Sales rep have high attrition, but may not be necessarily due to travel (Travel vs. Job Role shows other roles travel almost same as sales rep). This is most likely being caused by the younger population in this role.

```{r fig.height=3}
plotAttritionBars(data,'JobRole')  
plotBars(data,'JobRole','BusinessTravel')
plotTrend(data,'JobRole','Age')  
plotTrend(data,'Attrition','Age')  
```
