---
title: "Case Study 2: Atttrition Analysis"
author: "Nikhil Gupta"
date: "`r Sys.time()`"
output:
  html_document:
   toc: true
   toc_float: true
   toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(tibble)
library(dplyr)
library(tidyr)
library(ggplot2)
library(corrgram)
```

# Read and Preprocess

## Read Data
```{r }
data = read_xlsx("data/CaseStudy2-data.xlsx")
str(data)
summary(data)
```

## Format Data - Convert to factors
```{r}
data <- lapply( data, function(x) if (class(x) == 'character') as.factor(x) else x )
data <- as.data.frame(data)
str(data)
```

## Clean Data
```{r}
# TODO: Check for NA values 
```

## Drop unnecessary columns
```{r}
# Drop Over 18 since only 1 level
# Drop Employee Number and Count
drops = c('Over18','EmployeeCount','EmployeeNumber')
data = data[ , !(colnames(data) %in% drops)]
str(data)
```

# Exploratory Data Analysis

## Determine Highly Correlated Terms
```{r}
# corrgram(data,order = TRUE             # here you can pass the whole DF include non numeric fields, it will filter it for you.
#         , lower.panel = panel.shade # order = TRUE orders the matrix as per PCA
#          , upper.panel = panel.pie)  # panels indicate how you want upper and lower panels to look like

# remove all factor terms before doing correlation
data.num <- data[, sapply(data, class) != "factor"]
#str(data.num) # check

data.cor <- data.num %>% 
  as.matrix %>%
  cor %>%
  as.data.frame %>%
  rownames_to_column(var = 'var1') %>%
  gather(var2, value, -var1)

# Variables correlated by more than 0.7
arrange(filter(data.cor, value > .7 & value < 1), desc(value))

# Highlyy Correlated (remove ones that should be categorical): 
#   - MonthlyIncome (keep) <> JobLevel <> TotalWorkingYears
#   - PerformanceRating <> PercentSalaryHike (keep)
#   - YearsInCurrentRole <> YearsWithCurrManager <> YearsAtCompany (keep)

# TODO: Do scatter plots for explanatory variables that are continuous instead of just correlations for final presenatation.
```


## Visualizations and Statistical Analysis
```{r}
left = filter(data,Attrition == 'Yes')
stayed = filter(data,Attrition == 'No')

plot.stat.test = function(x,y){
  hist(x)
  hist(y)
  wilcox.test(x,y, correct=TRUE, alternative = 'two.sided', exact = TRUE, conf.int = TRUE, conf.level = 0.95)
}
```



### Monthly Income
```{r }
pl <- ggplot(data,aes(x=factor(Attrition),y=MonthlyIncome))
print(pl + geom_boxplot(aes(fill=factor(Attrition))))

ggplot(data = data, aes(x = TotalWorkingYears, y = MonthlyIncome, colour = Attrition)) + geom_point() + geom_smooth(method = 'lm', se = FALSE)

x = left$MonthlyIncome
y = stayed$MonthlyIncome
plot.stat.test(x,y)
# Highly Significant


pl <- ggplot(data,aes(x=factor(JobLevel),y=Age))
print(pl + geom_boxplot(aes(fill=factor(JobLevel))))



```

### Job Level
```{r}
pl <- ggplot(data,aes(x=factor(Attrition),y=JobLevel))
print(pl + geom_boxplot(aes(fill=factor(Attrition))))
x = left$JobLevel
y = stayed$JobLevel
plot.stat.test(x,y)
# Highly Significant

pl <- ggplot(data,aes(x=factor(Attrition)))
print(pl + geom_bar(aes(fill=factor(JobLevel))) + coord_flip())
# Very few people in Job Level 4 and 5 leave (find percentages by job level)
# These are managers, directors, Health Rep who seem to be paid more. - Check correspondingsalary graph form Max's Rmd file

pl <- ggplot(data,aes(x=factor(JobLevel),y=MonthlyIncome))
print(pl + geom_boxplot(aes(fill=factor(JobLevel))))

```


### Total Working Years
```{r}
pl <- ggplot(data,aes(x=factor(Attrition),y=TotalWorkingYears))
print(pl + geom_boxplot(aes(fill=factor(Attrition))))
x = left$TotalWorkingYears
y = stayed$TotalWorkingYears
plot.stat.test(x,y)
# Highly Significant
```

###  Performance Rating
```{r}
pl <- ggplot(data,aes(x=factor(Attrition),y=PerformanceRating))
print(pl + geom_boxplot(aes(fill=factor(Attrition))))
x = left$PerformanceRating
y = stayed$PerformanceRating
plot.stat.test(x,y)
# Not Statistically Different

pl <- ggplot(data,aes(x=factor(Attrition)))
print(pl + geom_bar(aes(fill=factor(PerformanceRating))) + coord_flip())
```

###  Percent Salary Hike
```{r}
pl <- ggplot(data,aes(x=factor(Attrition),y=PercentSalaryHike))
print(pl + geom_boxplot(aes(fill=factor(Attrition))))
x = left$PercentSalaryHike
y = stayed$PercentSalaryHike
plot.stat.test(x,y)
# Not Statistically Different
```

###  Years in Current Role
```{r}
pl <- ggplot(data,aes(x=factor(Attrition),y=YearsInCurrentRole))
print(pl + geom_boxplot(aes(fill=factor(Attrition))))
x = left$YearsInCurrentRole
y = stayed$YearsInCurrentRole
plot.stat.test(x,y)
# Statistically Different
```

###  Years with Current Manager
```{r}
pl <- ggplot(data,aes(x=factor(Attrition),y=YearsWithCurrManager))
print(pl + geom_boxplot(aes(fill=factor(Attrition))))
x = left$YearsWithCurrManager
y = stayed$YearsWithCurrManager
plot.stat.test(x,y)
# Statistically Different
```


###  Years at Company
```{r}
pl <- ggplot(data,aes(x=factor(Attrition),y=YearsAtCompany))
print(pl + geom_boxplot(aes(fill=factor(Attrition))))
x = left$YearsAtCompany
y = stayed$YearsAtCompany
plot.stat.test(x,y)
# Statistically Different
```

###  Business Travel
```{r}
pl <- ggplot(data,aes(x=factor(BusinessTravel)))
print(pl + geom_bar(aes(fill=factor(Attrition))))


pl <- ggplot(data,aes(x=factor(JobRole)))
print(pl + geom_bar(aes(fill=factor(BusinessTravel))) + coord_flip())

table(filter(data,JobRole == "Research Scientist")$BusinessTravel)
#Conclusion: TBD

```

### Department
```{r}
pl <- ggplot(data,aes(x=factor(Department)))
print(pl + geom_bar(aes(fill=factor(Attrition))))
#Conclusion: TBD
```

###  Education
```{r}
pl <- ggplot(data,aes(x=factor(Attrition),y=Education))
print(pl + geom_boxplot(aes(fill=factor(Attrition))))
x = left$Education
y = stayed$Education
plot.stat.test(x,y)
# Statistically Insignificant
```

###  Education Field
```{r}
pl <- ggplot(data,aes(x=factor(EducationField)))
print(pl + geom_bar(aes(fill=factor(Attrition))) + coord_flip())
#Conclusion: TBD
```
###  Environment Satisfaction
```{r}
pl <- ggplot(data,aes(x=factor(Attrition),y=EnvironmentSatisfaction))
print(pl + geom_boxplot(aes(fill=factor(Attrition))))
x = left$EnvironmentSatisfaction
y = stayed$EnvironmentSatisfaction
plot.stat.test(x,y)
# Statistically Significant
```

###  Gender
```{r}
pl <- ggplot(data,aes(x=factor(Gender)))
print(pl + geom_bar(aes(fill=factor(Attrition))))
#Conclusion: TBD
```

### Hourly Rate
```{r}
pl <- ggplot(data,aes(x=factor(Attrition),y=HourlyRate))
print(pl + geom_boxplot(aes(fill=factor(Attrition))))
x = left$HourlyRate
y = stayed$HourlyRate
plot.stat.test(x,y)
# Statistically Insignificant
```

### Job Involvement
```{r}
pl <- ggplot(data,aes(x=factor(Attrition),y=JobInvolvement))
print(pl + geom_boxplot(aes(fill=factor(Attrition))))
x = left$JobInvolvement
y = stayed$JobInvolvement
plot.stat.test(x,y)
# Statistically Insignificant
```

### Job Role
```{r}
pl <- ggplot(data,aes(x=factor(JobRole)))
print(pl + geom_bar(aes(fill=factor(Attrition))) + coord_flip())
#Conclusion: Lab Technitian seems to have highest attrition rate
print(pl + geom_bar(aes(fill=factor(OverTime))) + coord_flip()) # not necessarily due to overtime
print(pl + geom_bar(aes(fill=factor(JobLevel))) + coord_flip())

# Check distribution vs. Age to see if that is a confounding variable
pl <- ggplot(data,aes(x=factor(JobRole),y=Age))
print(pl + geom_boxplot(aes(fill=factor(JobRole))) + coord_flip()) 

pl <- ggplot(data,aes(x=factor(JobRole)))
print(pl + geom_bar(aes(y = (..count..)/sum(..count..), fill=factor(JobLevel))) + coord_flip())

```

### Marital Status
```{r}
pl <- ggplot(data,aes(x=factor(MaritalStatus)))
print(pl + geom_bar(aes(fill=factor(Attrition))))
#Conclusion: Single people leave at a higher rate (as they are more flexible)
```

### Monthly Rate
```{r}
pl <- ggplot(data,aes(x=factor(Attrition),y=MonthlyRate))
print(pl + geom_boxplot(aes(fill=factor(Attrition))))
x = left$MonthlyRate
y = stayed$MonthlyRate
plot.stat.test(x,y)
# Statistically Insignificant
```

### Number of Companies Worked
```{r}
pl <- ggplot(data,aes(x=factor(Attrition),y=NumCompaniesWorked))
print(pl + geom_boxplot(aes(fill=factor(Attrition))))
x = left$NumCompaniesWorked
y = stayed$NumCompaniesWorked
plot.stat.test(x,y)
# Statistically Insignificant
```

### Overtime
```{r}
pl <- ggplot(data,aes(x=factor(OverTime)))
print(pl + geom_bar(aes(fill=factor(Attrition))))
#Conclusion: Those who get overtime leave at a higher rate (Strange -- check out other confounding factors that may impact this)
#   - which department do they come from
#   - years of experience
#   - stock

pl <- ggplot(data,aes(x=factor(OverTime),y=JobLevel))
print(pl + geom_boxplot(aes(fill=factor(OverTime))))

pl <- ggplot(data,aes(x=factor(JobRole)))
print(pl + geom_bar(aes(y = (..count..)/sum(..count..), fill=factor(OverTime))) + coord_flip())

pl <- ggplot(data,aes(x=factor(OverTime),y=Age))
print(pl + geom_boxplot(aes(fill=factor(OverTime))))

# Burnout could be a major factor for those who do overtime.


```

### Job Satisfaction
```{r}
pl <- ggplot(data,aes(x=factor(JobSatisfaction),y=Age))
print(pl + geom_boxplot(aes(fill=factor(JobSatisfaction))))

# Max's survey plot may be better
```

# Future Steps
## For team to look at

```{r Convert required variables from Numeric to Categorical}
summary(data)
# More Cleanup needed
# TODO: Following should be factors: 
#       Education (5 levels), EnvironmentSatisfaction (4 levels), JobInvolvement (4 levels), JobLevel (5 levels), JobSatisfaction (4 levels), 
#       PerformanceRating (2 levels), RelationshipSatisfaction (4 levels), StockOptionLevel (4 levels), WorkLifeBalance (4 levels)
# TODO: Too many levels (curse of dimentionality), consider combining some levels 
#       Check co-relation between independent variabels. Remove the highly co-related varaiables  

# What is Diff between Monthly Income and Monthly Rate -- Not Important
# TODO: Add visualizations for stock options and other missing variables
# TODO: Add scatter plots to get trends (coded by Attrition)
# TODO: For 4 point scale, probably we dont need Wilcox Test - 2 sample t-test would be sufficient and more powerful
# TODO: Correlation Heatmap
```


## For Final Submission
```{r Model}
# Response is a categorical Variable
# Try K-Means Clustering (2 clusters), but this only helps with prediction, not in telling us what 3 variables are important 
# Other option is to try logistic regression, but we have not studied this. Is this somethig that would be expected.

# Rething below model 
# lm.model = lm(Attrition ~ . , data = data)
# lm.model

# TODO: (Nikhil) t-tests for 2 groups (Attrition = yes and No) for Salary Variable, Hike Variable, etc. See which ones are statistically significant.
# TODO: (Nikhil) knn with different variables to see which perform best (Confusion Matrix). It might tell which variables are most important in predicting attrition
#       https://arxiv.org/ftp/arxiv/papers/1806/1806.10480.pdf
#       https://arxiv.org/pdf/1707.01377.pdf
# TODO: (??) Random Forest with different variables
# TODO: SVM (James) -- You can create a visual represenation of the parameters. It can show visually the difference between the classes
# TODO: XGBoost (Max)
```