---
title: "CaseStady2_Max"
author: "Max Moro"
date: "November 26, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(openxlsx)
library(reshape2)
library(forcats)
library(scales)
library(xgboost)
library(caret)

# MonthlyIncom

# Tenure
  # Years at company
  # Age
  # YearsInCurrentLog
  # TotalWorkingYers
  
# Satisfaction
  #job Satisfaction
  #Environemtn Statisfaciotn

# box  Plot Age and Job Level

# charts categorical
  #Businserr TRavle
    # across all jobs.. 
  #Overtime (burnout) - no correlation with age and job level (boxplot )
    # Boxplot overtime and age
    # Boxplot overtime and JobLevel
  #Job Role



```

# Reading  Data Source

```{r read}
dataOri=read.xlsx('data/CaseStudy2-data.xlsx',sheet=1) %>%
  mutate(EmployeeNumber = as.character(EmployeeNumber))

```

# Transformation
```{r transformation}
#checking for na - No NAs in the columns
which(sapply(dataOri,function(x){any(is.na(x))}))
#check for columns with one factor -> 3 columns
oneElement = names(which(sapply(dataOri,function(x){n_distinct(x)==1})))
print(oneElement)
data=dataOri %>% select(-!!oneElement)
```

# Correlation Analysis

## Numerical Columns

### Distribution of Numeric Columns

```{r  fig.height=7, fig.width=8}
#find group with statistical difference

colNum = names(which(sapply(data,is.numeric)))

temp=reshape2::melt(data,id.vars='Attrition',measure.vars=colNum) %>%
  group_by(variable) %>%
  mutate(density = dnorm(value, mean(value), sd(value))) %>% ungroup()

ggplot(data=temp,aes(x=value))+
  geom_histogram(aes(y=..density..),color='gray',bins = 20) +
  geom_line(aes(y = density),   colour = "red") +
  #stat_function(fun = dnorm, args = list(mean = mean(..count..), sd = sd(..count..)),color='red')+
  facet_wrap(~variable,scales='free',ncol=4) +
  theme_minimal()
```

### Applying Log to Some Variables

Following Variable have been transformed with Log function to fit a normal distribution.

```{r  fig.height=5, fig.width=8}

#colsToLog = c('DistanceFromHome','MonthlyIncome','PercentSalaryHike','TotalWorkingYears','YearsAtCompany','YearsInCurrentRole'            ,'YearsSinceLastPromotion','YearsWithCurrManager')
# ggplot(data=filter(temp,variable %in% colsToLog),aes(x=value))+
#   geom_histogram(aes(y=..density..),fill='gray',bins = 20) +
#   facet_wrap(~variable,scales='free',ncol=4) +
#   ggplot2::ggtitle('Original Values')+
#   theme_minimal()


#Testing the Log
colsToLog = c('MonthlyIncome','TotalWorkingYears','YearsAtCompany','YearsInCurrentRole','YearsSinceLastPromotion','YearsWithCurrManager')

temp$log = log(temp$value +.01)
#plot of logged values
ggplot(data=filter(temp,variable %in% colsToLog),aes(x=log))+
  geom_histogram(aes(y=..density..),fill='blue',bins = 20) +
  facet_wrap(~variable,scales='free',ncol=4) +
  ggplot2::ggtitle('Logged Values')+
  theme_minimal()

#Applying the log to the main dataframe
colsLog = paste0(colsToLog,'_log')
dataLog=mutate_at(data,colsToLog,.funs=funs(log=log(. + 0.01))) %>% select(-colsToLog)

```

### Variable with different mean across Attrition 

```{r}
ttestOut=as.character()
colNum = names(which(sapply(dataLog,is.numeric)))
for (col in colNum){
  test=t.test(dataLog[dataLog$Attrition=='Yes',col],dataLog[dataLog$Attrition=='No',col],alternative = 'two.sided',var.equal=F)
  if(test$p.value<0.01) ttestOut = c(ttestOut,col)
}

message('following variables have a different mean between Attrition and Non-Attrition')
temp=reshape2::melt(dataLog,id.vars='Attrition',measure.vars=ttestOut)
ggplot(data=temp,aes(x=Attrition,y=value))+
  geom_boxplot() +
  facet_wrap(~variable,scales='free',ncol=4) +
  theme_minimal()
```

## Categorical Variables

```{r  fig.height=5, fig.width=10}
#find group with statistical difference
colText = names(which(sapply(select(data,-Attrition,-EmployeeNumber),is.character)))

#make the datast vertical to better manage and chart. Also calculating percentages
temp=reshape2::melt(data,id.vars='Attrition',measure.vars=colText)  %>%
  group_by(variable,value,Attrition) %>%
  summarise(count=n()) %>%
  group_by(variable,value) %>%
  mutate(percent=count/sum(count) ) %>%
  filter(Attrition=='Yes') %>%
  arrange(percent)

#ggplot sort breaks if we have same value across facets.
#to help gplot sort we need to the value+variable insetead of value if there are more than one acorss variables
temp = temp %>% group_by(value)  %>% mutate(cnt = n_distinct(variable)) %>%
  ungroup() %>% mutate(name=ifelse(cnt>1,paste0(value,' (',(substr(variable,1,1)),')'),value))

temp$name=forcats::fct_reorder(temp$name,temp$percent )

message("Distribution of Attrition = Yes, across categorical values")

ggplot(data=temp,aes(x=name,y=percent))+
  geom_bar(stat='identity',fill='LightBlue') +
  coord_flip()+
  geom_text(aes(label=percent(percent)),size=3,hjust=-.1,color='#555555')+
  facet_wrap(~variable,scales='free_y',ncol=3) +
  ggplot2::scale_y_continuous(labels=percent_format(accuracy = 1),limits=c(0,max(temp$percent+.1)),name='Attrition Rate') +
  ggplot2::scale_x_discrete(name='Group') +
  theme_minimal()
```


## Importance of Variables

Using XGBoost model 

```{r}
set.seed(1701)

grid_default <- expand.grid(nrounds = 100
                            ,max_depth = 6
                            ,eta = .3
                            ,gamma = 0
                            ,colsample_bytree = 1
                            ,min_child_weight = 1
                            ,subsample = 1
)

train_control <- caret::trainControl(method = "none",verboseIter = FALSE,allowParallel = TRUE)
data2=data  %>% select(-EmployeeNumber)

xgb <- caret::train(Attrition ~ . ,data=data2
                    ,trControl = train_control,tuneGrid = grid_default
                    ,method = "xgbTree",verbose = TRUE
)

plot(caret::varImp(xgb),top=25)

```

# Question 1
  
  Identify (at least) the top three factors that contribute to turnover.
  
# Trends

```{r eval=FALSE, fig.height=15, include=FALSE}
#This code is only used to discover trends, manually. Not needed for the final report
colNum = names(which(sapply(data,is.numeric)))
facts = c('BusinessTravel','Department','EducationField','Gender','JobRole','MaritalStatus','OverTime')

temp=reshape2::melt(data,id.vars=facts,measure.vars=colNum) 
fact=facts[2]
for (fact in facts){
  ggplot2::ggplot(data=temp,aes_string(x=fact,y='value')) +
    geom_boxplot()  +
    facet_wrap(~variable,scales = 'free',ncol=3) +
    coord_flip() + 
    ggtitle(fact) + 
    theme_minimal()
}

```

```{r}

plotTrend = function(data,x,y){
  #function to create a boxplot comparing categorical and numeric columns
  data2=data 
  data2$x=forcats::fct_reorder(data[[x]],data[[y]],.fun=median)
  data2$y=data2[[y]]
  
  ggplot2::ggplot(data=data2,aes(x=x,y=y)) +
    geom_boxplot(aes(fill=x)) +
    coord_flip() +
    xlab(x)+
    ylab(y)+
    ggplot2::ggtitle(paste('Comparing',y,'and',x)) +
    theme_minimal() +
    scale_y_continuous(labels=comma_format(accuracy=1)) + 
    scale_fill_brewer(palette='Blues',guide=F)
}

plotSurvey = function(data,x,y){
  #function to create a text plot comparing survey data
  data2=data 
  data2$x=forcats::fct_reorder(data[[x]],data[[y]],.fun=median)
  data2$y=as.character(data2[[y]])
  data2 = data2 %>% group_by(x,y)  %>%
    summarise(count=n()) %>%
    group_by(x) %>%
    mutate(perc=count/sum(count)) %>%
    ungroup() %>%
    mutate(color = ifelse(y>=3,-perc,perc))
  
  ggplot2::ggplot(data=data2,aes(x=x,y=y,size=perc,label=percent(perc,1))) +
    geom_text(aes(color=color)) +
    coord_flip() +
    xlab(x)+
    ylab(y)+
    ggplot2::ggtitle(paste('Comparing',y,'and',x)) +
    theme_minimal() + 
    scale_size_continuous(guide=F) +
    #scale_color_continuous(guide=F,low='#fcbba1',high='#67000d')
    scale_color_gradient2(guide=F,low='#006d2c',mid='#ffffff',high='#67000d',midpoint=0)
}

```


## Marketing Degree is the most paid group, HR Degree is the least
```{r}
plotTrend(data,'EducationField','MonthlyIncome')
```

## Manager are the most Senior poeple, Sales Repr are the youngest
```{r}
plotTrend(data,'JobRole','Age')
```



## Manager works closer to work than Sales executive
```{r}
plotTrend(data,'JobRole','DistanceFromHome')
```



## Manager Monthly Rate is higher
```{r}
plotTrend(data,'JobRole','MonthlyRate')
```


## Manager Earns vastly more than other groups, Sales Repr have the lowest Monthly Income salary
```{r}
plotTrend(data,'JobRole','MonthlyIncome')
```



## Manager Monthly Rate is higher
```{r}
plotTrend(data,'JobRole','YearsAtCompany')
```

## HR has the lowest satisfaction
```{r}
plotSurvey(data,'JobRole','JobSatisfaction')
```